---
title: "ENDIG: clean data"
author: "Nils Tjaden"
date: "`r Sys.Date()`"
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../documentation") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# move knitr working directory up by one level so it's the same as
# the on of the R Project 
knitr::opts_knit$set(root.dir = '../') 
```

# Packages

Load necessary packages:

-   `readxl` tidyverse package for reading data from Office Open XML spreadsheet files
-   `dplyr` for general-purpose data-wrangling
-   `tidyr` for general-purpose data-wrangling
-   `stringr` for manipulating strings
-   `sf` for handling of spatial polygon data

```{r, message=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(sf)
```

# Constants

## Disease meta data

Load the main meta data file:

```{r}
load(file = file.path("./data/processed/meta.rds"))
head(meta)
```

Read list of table sheets that require special processing due to differing format etc.:

```{r}
special_processing_list <- read.csv("./data/special_processing_required.csv")
special_processing_list <- paste(special_processing_list$year,
                                  special_processing_list$dis)
special_processing_list
```

## EU countries

List of all EU (and non-EU) countries appearing in the data:

```{r}
load(file = file.path("./data/processed/countries.rds"))
countries
```

Note that these are 31 instead of the expected 27 countries. The list additionally includes: Iceland, Liechtenstein, Norway, United Kingdom. As of 2015, the UK is still on the list for Creutzfeldt-Jacob Disease (VCJD), but does not report anything else to ECDC.

## Country polygons

Load previously downloaded country polygons from NaturalEarth:

```{r}
load(file = file.path("./data/processed/countries_sf.rds"))
```

# Output data

Initialize the output data structure. We will create two separate list objects for plotting the maps and heatmaps, respectively. Each of them contain another set of lists for the different surveillance system characteristics.

```{r}
data_spatial <- list(compulsory=list(), comprehensive=list(), active=list(), aggregated=list())
data_heatmap <- list(compulsory=list(), comprehensive=list(), active=list(), aggregated=list())
```

# Functions

```{r}
process_year <- function(year){
  message("+++ processing year ", year, " +++")
  disease_IDs <- meta %>% select(ID_master, paste0("ID_ECDC_", year))
  
  for (i in 1:nrow(disease_IDs)){
    dis_global <- as.character(disease_IDs[i, "ID_master"])
    dis_local <- as.character(disease_IDs[i, 2])
    # if there is no table sheet for this disease in this year's workbook,
    # just create an empty data.frame
    if(is.na(dis_local)){
      dat <- data.frame()
    }
    # otherwise prepare the data
    else{
      process_disease(year, dis_local)
    }
  }
}


process_disease <- function(year, dis){
  # check if special processing is required
  if(table_OK(year, dis)){
    dat <- read_sheet(year, dis)
  } else {
    message(paste("skipping non-standard table for", dis, "in", year))
  }
 
  dat <- clean_duplicates(dat, year, dis)
  
   
}


clean_duplicates <- function(dat, year, dis){
  dupe_countries <- unique(dat$Country[which(duplicated(dat$Country))])
  if(length(dupe_countries)>0){
    message(paste("duplicate country rows found in", dis, "sheet:"))
    for(dupe_country in dupe_countries){
      cat("\ninput data:\n")
      dupe_rows <- which(dat$Country %in% dupe_country)
      sub <- dat[dupe_rows, ]
      print(sub)
      
      # Start with the assumption that rows for this country disagree in all 
      # columns, setting all columns to "." for unspecified.
      out <- c(dupe_country, rep(".", times=4))
      
      # if all rows of a column are identical, overwrite that spot in the
      # output row with the corresponding value (catching it from the first row of sub)
      if(length(unique(sub[,2]))==1){out[2] <- sub[1,2]}
      if(length(unique(sub[,3]))==1){out[3] <- sub[1,3]}
      if(length(unique(sub[,4]))==1){out[4] <- sub[1,4]}
      if(length(unique(sub[,5]))==1){out[5] <- sub[1,5]}
      
      # overwrite the first row for this country with the new data
      dat[dupe_rows[1],] <- out
      # remove the remaining rows for this country
      dat <- dat[-dupe_rows[2:length(dupe_rows)],]
      
      cat("\nturns to:\n")
      print(dat[dupe_rows[1],])
    }
  }
  return(dat)
}








# read a disease table sheet that follows the usual layout
read_sheet <- function(year, dis){
  # read the sheet, skipping the top 2 rows and manually assigning
  # column names
  dat <-  readxl::read_xlsx(path = file.path(getwd(),
                                             paste0("data/raw/Table-surveillance_systems_overview_", year, ".xlsx")),
                            skip = 2,
                            col_types = "text",
                            col_names = c("Country",
                                          "src",
                                          "compu",
                                          "compre",
                                          "act",
                                          "agg",
                                          "L",
                                          "P",
                                          "H",
                                          "O",
                                          "casedef"),
                            sheet = dis)
  if(nrow(dat) == 0){#if the sheet exists, but is empty, create a dummy
    dat <- data.frame(Country="dummy",
                      compu="dummy",
                      compre="dummy",
                      act="dummy",
                      agg="dummy")
  }
  # select relevant columns
  dat <- dat %>% select(Country, compu, compre, act, agg)
  
  # ECDC switched from using "Czech Republic" in earlier tables
  # to using "Czechia" at some point.
  # Harmonize that to "Czech Republic", if necessary
  dat$Country[which(dat$Country %in% "Czechia")] <- "Czech Republic"
  
  # some sheets have non-data rows at the bottom. Keep only rows that 
  # have a known country name in the first column. This also removes 
  # the dummy row created for empty tables.
  dat <- dat[which(dat$Country %in% countries),]
  return(data.frame(dat))
}



# check if a certain combination of year and disease ID is on 
# the list of table sheets needing non-standard processing
table_OK <- function(year, dis){
  !paste(year, dis) %in% special_processing_list
}
```

```{r}

```
